A Speculative Approach to Parallelization in Particle Swarm Optimization

### 2011-04-01 masters_thesis/si_paper/paper.tex
More changes from feedback.  Getting closer

In our approach, we refactor PSO such that the computation needed for iteration
$t+1$ can be done concurrently with the computation needed for iteration $t$.
Thus we can perform two iterations of PSO at once.  Even with some amount of
wasted computation, we show that this approach to parallelization in PSO often
outperforms the naive parallelization of simply adding particles to the swarm.
SEPSO produces results that are exactly equivalent to PSO; this is not a new
algorithm or variant, only a new method of parallelization.

@@

Furthermore, we show that if we relax the requirements of the algorithm, no
longer demanding that it strictly reproduce the exact behavior of standard PSO,
we can introduce new speculative techniques that often out-perform both naive
parallelizations of PSO and SEPSO.  These relaxations make better use of the
information obtained from the extra exploration made by the speculative
function evaluations.  We also explore the idea that, like branch prediction in
processors, we need not speculatively evaluate \emph{all} possible future
positions, we can accelerate the algorithm even if we are just \emph{likely} to
have guessed right.  By pruning the speculation to just paths that are
statistically likely to reproduce the paths that are equivalent to PSO we can
increase the swarm size without increasing the number of speculative
evaluations.  We also consider several recovery strategies for cases where the
pruned set of speculative evaluations does not contain the evaluation that
standard PSO would have done.  A further improvement we explore is speculating
several iterations ahead instead of just one, which is made possible by pruning
the number of speculative evaluations.

@@

While the idea of speculative evaluation in particle swarm optimization is
intriguing, the results in \sec{results} show that in many cases it requires
too many extra evaluations to produce competitive results.  On most functions,
the larger swarm size that can be used in naive parallelizations leads to
better performance than is found by trying to do speculative evaluation to
exactly reproduce PSO.  However, if we keep the idea of speculative evaluation
while relaxing the requirement of exactly reproducing the behavior of the
original PSO algorithm, we see some impressive results.

@@

Because the problem facing speculative evaluation is having too many possible
evaluations to perform, a natural step to take is to eliminate some of the
evaluations.  If we could reliably predict which branch were going to be taken,
we could limit ourselves to one speculative evaluation per particle instead of
$2n+1$.  With a fixed number of processors, this would allow us to greatly
increase the swarm size relative to that needed in the original speculative
algorithm (e.g., with 120 processors, a naive parallelization has a swarm of
size 120, complete speculative evaluation has a swarm of size 15, and pruning
the evaluations to only one per particle allows a swarm of size 60).  As not
all of the branches are evaluated in any given iteration, we call this
technique pruning.

@@

We ran experiments to compare our speculative parallelization of PSO to the
naive parallelization.  At each iteration of the algorithms, we use one
processor to perform one function evaluation for one particle, be it
speculative or non-speculative.  The speculative version actually performs two
iterations of PSO at each ``iteration,'' so we instead call each ``iteration''
a ``time step.''  As noted in the introduction, this is a reasonable
architecture to use if the function being optimized takes longer than around
100 milliseconds to run.

@@

Using the same number of processors for each approach requires that our
speculative parallelizations have a smaller swarm size than the naive
parallelization.  For each of the topologies we used, a particle has three
neighbors including itself.  As shown in Table~\ref{tab:evals}, this results in
$7$ speculative evaluations per particle.  With one evaluation needed for the
original, non-speculative particle, we have $8p$ evaluations for every two
iterations, where $p$ is the number of particles in the speculative swarm.  The
extra evaluations required in our speculative approach would instead be used to
evaluate particles in naive parallelizations, so we compare swarms of size
$p$ in speculative evaluation with swarms of size $8p$ in naive approaches.

@@

That being said, given the time required for the evaluation of the objective
function and the communication overhead per iteration for a specific
implementation, the actual running time can be recovered from all of our
results.  Simply multiply the number of time steps by the sum of the function
evaluation time and the overhead.  We wish to stress that the ``time steps'' we
report are \emph{exactly} proportional to wall-clock time, given the
architecture we have assumed.

@@

The parallelization techniques we compare are the naive parallelization (here
labeled Naive simply for ease of reference), our original speculative
approach (labeled SEPSO), and the four relaxations of SEPSO discussed in
\sec{relax}.  In presenting our results, we call the approach developed in
\sec{pickbest} Pick Best.  The methods described in \sec{pruning} through
\sec{wrong} are called Pick Best Pruned and Social Promotion Pruned, and the
method in \sec{manyiters} we call Many Iterations.

@@

Because pruning only requires two speculative evaluations per particle (along
with evaluating the original particle), we can use swarms of size
$\frac{1}{3}p$ to compare to a swarm of size $p$ with the naive
parallelization, instead of $\frac{1}{8}p$ with other techniques.

@@

Where the Complete topology would normally be preferred, we use a Random
topology in SEPSO and Pick Best, as Complete leads to an explosion in the
number of speculative evaluations (as noted, the other speculative techniques
can still use a Complete topology; we often report results on both topologies
for those methods).  If speculative evaluation were not being performed, it is
possible that the Complete topology would be used.  However, the Complete
topology also requires a very large amount of interprocessor communication in
distributed PSO, so it is still quite possible that Random would be used even
with naive parallelizations~\cite{mcnabb-2009-large-particle-swarms}.  But,
to be fair in our comparisons, we compare to the naive parallelization using
both the Random topology and the Complete topology (labeled PSO Random and PSO
Complete in our results).  Again, the amount of communication overhead is
heavily dependent on implementation details which are not the focus of this
paper.  A practitioner using a particular implementation can compare results
for whichever topology is most practical given the specific implementation.

@@

Rastrigin is a multi-modal function that is best optimized with a large,
Complete swarm.  It has been shown that with Rastrigin, the more particles
there are in the swarm, the lower function value it finds, up to at least 4000
particles~\citep{mcnabb-2009-large-particle-swarms}.  Smaller swarms get caught
in local optima.  Because our speculative algorithms require significantly
smaller swarm sizes, we would expect to not perform very well on functions such
as Rastrigin.  Our experiments show our intuition to be correct.  In this
experiment we used 240 processors, so SEPSO and Pick Best each had 30
particles, and the Naive algorithms had 240 particles.  As expected, SEPSO and
Pick Best converge quickly to worse local optima than Naive does.
\fig{rastrigin} shows the results graphically, and \tabref{rastrigin-20} shows
results for all of the methods we tried.

\begin{table}
  \caption{Summary of results for function Rastrigin with 20 dimensions,
  measuring number of time steps to reach a value of 20}
  \label{tab:rastrigin-20}
  \centering
  \begin{tabular}{|l|r|r|r|}
  \hline
  Algorithm&\% Complete&Mean&St. Dev.\\
  \hline
  \hline
  Naive Ring&25\%&743.4&212.0\\
  \hline
  Naive Random&100\%&372.1&86.6\\
  \hline
  \textbf{Naive Complete}&\textbf{95\%}&\textbf{273.1}&\textbf{99.7}\\
  \hline
  SEPSO Ring&5\%&471.0&0.0\\
  \hline
  SEPSO Random&15\%&208.0&17.0\\
  \hline
  Pick Best Ring&35\%&489.0&152.0\\
  \hline
  Pick Best Random&15\%&136.0&40.1\\
  \hline
  Pick Best Pruned Ring&95\%&545.9&208.5\\
  \hline
  Pick Best Pruned Random&10\%&95.0&11.0\\
  \hline
  Pick Best Pruned Complete&5\%&59.0&0.0\\
  \hline
  Social Promotion Pruned Ring&5\%&580.0&0.0\\
  \hline
  Social Promotion Pruned Random&80\%&278.2&136.3\\
  \hline
  Social Promotion Pruned Complete&50\%&200.5&130.2\\
  \hline
  Many Iterations Ring&10\%&117.0&25.0\\
  \hline
  Many Iterations Random&5\%&57.0&0.0\\
  \hline
  Many Iterations Complete&0\%&N/A&N/A\\
  \hline
  \end{tabular}
\end{table}

@@

We ran 50 trials of each experiment with Griewank, so that the curves are more
smooth.  We show results in \fig{basic-griewank1} for swarms of size 30 and 240
using the Ring topology.  One can see in the figure that the SEPSO approach
finds the optimum faster than the Naive approach.  However, because the swarm
size is so small, SEPSO gets stuck a little less than half of the time.

@@

Because the speculative algorithm got stuck with such a small swarm size, we
ran a set of experiments on Griewank with 800 processors, giving swarms of size
100 and 800 instead of 30 and 240.  \fig{basic-griewank2} shows the results.
\fig{basic-griewank2} is very similar to \fig{basic-griewank1}, except that
speculative evaluation finds the global optimum essentially 100\% of the time.
Thus with a few more processors, our algorithm finds the optimum on average
twice as fast as the Naive approach without a significant loss of accuracy.

@@

With Griewank, the premature convergence problems inherent in picking the best
child are exacerbated when speculating several iterations ahead.  When Many
Iterations finds the optimum, it finds it quicker than any other method we
tried, on average four times faster than Naive.  However, it also gets stuck
and fails to find the optimum more than any other method.  The results are
shown in \fig{griewank-manyiters}.   \fig{griewank-manyiters} is also
interesting in that it highlights the trade-off between accuracy and speed in
the various approaches at this swarm size.  The faster the approach finds the
optimum, the less likely it is to be successful.

@@

\begin{table}
  \caption{Summary of results for function Sphere with 20 dimensions,
  measuring number of time steps to reach a value of 1e-35}
  \label{tab:sphere-20}
  \centering
  \begin{tabular}{|l|r|r|r|}
  \hline
  Algorithm&\% Complete&Mean&St. Dev.\\
  \hline
  \hline
  Naive Ring&0\%&N/A&N/A\\
  \hline
  Naive Random&0\%&N/A&N/A\\
  \hline
  Naive Complete&100\%&931.0&19.1\\
  \hline
  SEPSO Ring&0\%&N/A&N/A\\
  \hline
  SEPSO Random&85\%&972.8&15.8\\
  \hline
  Pick Best Ring&100\%&917.4&21.0\\
  \hline
  Pick Best Random&100\%&768.1&14.3\\
  \hline
  Pick Best Pruned Ring&100\%&967.8&9.9\\
  \hline
  Pick Best Pruned Random&100\%&693.5&10.4\\
  \hline
  \textbf{Pick Best Pruned Complete}&\textbf{100\%}&\textbf{389.6}&
    \textbf{9.6}\\
  \hline
  Social Promotion Pruned Ring&0\%&N/A&N/A\\
  \hline
  Social Promotion Pruned Random&100\%&971.1&10.6\\
  \hline
  Social Promotion Pruned Complete&100\%&777.8&16.7\\
  \hline
  Many Iterations Ring&100\%&575.2&9.8\\
  \hline
  Many Iterations Random&100\%&442.4&7.1\\
  \hline
  Many Iterations Complete&0\%&N/A&N/A\\
  \hline
  \end{tabular}
\end{table}

@@

Schwefel 2.21 is a function similar to Sphere, but Schwefel 2.21 is benefited
more by larger swarms than Sphere is.  Thus our speculative algorithms often
fail outperform Naive with 240 processors because the simple speculative
methods only have 30 particles.  As we will show later, when we use 800
processors at 50 dimensions, our methods perform much better.  However, the
pruned versions of our methods have 80 particles and thus are able to
outperform Naive even with only 240 processors.  Because the graph of Schwefel
2.21 looks very similar to that of Sphere, we present all of our results in
\tabref{schwefel-20}.

\begin{table}
  \caption{Summary of results for function Schwefel with 20 dimensions,
  measuring number of time steps to reach a value of 1e-06}
  \label{tab:schwefel-20}
  \centering
  \begin{tabular}{|l|r|r|r|}
  \hline
  Algorithm&\% Complete&Mean&St. Dev.\\
  \hline
  \hline
  Naive Ring&0\%&N/A&N/A\\
  \hline
  Naive Random&0\%&N/A&N/A\\
  \hline
  Naive Complete&100\%&837.2&44.8\\
  \hline
  SEPSO Random&0\%&N/A&N/A\\
  \hline
  Pick Best Random&5\%&938.0&0.0\\
  \hline
  \textbf{Pick Best Pruned Random}&\textbf{100\%}&\textbf{639.9}&\textbf{29.4}\\
  \hline
  \textbf{Pick Best Pruned Complete}&\textbf{100\%}&\textbf{597.7}&
	\textbf{77.9}\\
  \hline
  Social Promotion Pruned Random&75\%&938.0&33.0\\
  \hline
  Social Promotion Pruned Complete&100\%&815.2&43.5\\
  \hline
  Many Iterations Random&100\%&783.5&94.5\\
  \hline
  Many Iterations Complete&0\%&N/A&N/A\\
  \hline
  \end{tabular}
\end{table}

@@

Bohachevsky is a unimodal function best optimized with a Complete swarm.  It is
similar to Griewank in that there is a global optimum with a value of 0, and
the swarm sometimes finds it and sometimes does not.  Thus we present a graph
similar to those of Griewank, because ``average value'' graphs have a
misleading flat line.  In \fig{bohachevsky} we show a plot with a few pertinent
methods, while \tabref{bohachevsky-20} shows all of the results.  All of our
speculative approaches found the optimum much quicker than Naive with a Random
topology.  However, SEPSO was slower than Naive Complete and got stuck 25\% of
the time.  Pick Best, Pick Best Pruned, and Many Iterations all outperformed
Naive Complete, with Many Iterations finding the optimum about twice as fast.

@@

\begin{table}
  \caption{Summary of results for function Bohachevsky with 20 dimensions,
  measuring number of time steps to reach a value of 1e-06.}
  \label{tab:bohachevsky-20}
  \centering
  \begin{tabular}{|l|r|r|r|}
  \hline
  Algorithm&\% Complete&Mean&St. Dev.\\
  \hline
  \hline
  Naive Random&100\%&472.3&6.6\\
  \hline
  Naive Complete&100\%&238.2&11.2\\
  \hline
  SEPSO Random&75\%&261.6&9.2\\
  \hline
  Pick Best Random&100\%&211.8&9.2\\
  \hline
  Pick Best Pruned Random&100\%&189.7&3.4\\
  \hline
  Pick Best Pruned Complete&FIXME!&FIXME!&FIXME!\\
  \hline
  Social Promotion Pruned Random&100\%&260.3&6.5\\
  \hline
  Social Promotion Pruned Complete&FIXME!&FIXME!&FIXME!\\
  \hline
  \textbf{Many Iterations Random}&\textbf{94\%}&\textbf{118.2}&\textbf{2.6}\\
  \hline
  Many Iterations Complete&0\%&N/A&N/A\\
  \hline
  \end{tabular}
\end{table}

@@

Given our observation from the 20 dimensional benchmark functions that Many
Iterations most often outperforms other speculative approaches, we only present
results for Many Iterations and Naive for the 50 and 500 dimensional
variants.  For these experiments we used 800 processors, as the problems are
more difficult and speculative approaches perform best when many processors are
available, as we showed in \sec{griewank-20}.  Many Iterations Complete showed
the same premature convergence that was observed with the 20 dimensional
benchmark functions, so we do not include those results in our tables; the
success rate was 0\% for all functions.

Tables~\ref{tab:rastrigin-50} through~\ref{tab:bohachevsky-50} present
summaries of our results for each of the benchmark functions. As with the 20
dimensional variants, our speculative methods did not perform well on
Rastrigin, though for all other benchmarks Many Iterations significantly
outperformed Naive.  In each table, we bold the best method, meaning that it
has at least a 90\% success rate and its average time to completion is faster
than all other methods that have at least a 90\% success rate.  For example,
Many Iterations Random, when it is successful, achieves the required value 6
times faster than Naive Random.  However, because Many Iterations Random only
has a 25\% success rate while Naive Random has a 100\% success rate, Naive
Random is bolded.

@@

A summary of the results is shown in \tabref{50dims-summary}.  For functions
where Many Iterations outperformed Naive, we report the average speed up.  On
all functions except Rastrigin, our methods showed an average speed up of from
two to six times compared to previous methods.

\begin{table}
  \caption{Summary of results for function Rastrigin with 50 dimensions,
  measuring number of time steps to reach a value of 100}
  \label{tab:rastrigin-50}
  \centering
  \begin{tabular}{|l|r|r|r|}
  \hline
  Algorithm&\% Complete&Mean&St. Dev.\\
  \hline
  \hline
  Naive Ring&5\%&776.0&0.0\\
  \hline
  \textbf{Naive Random}&\textbf{100\%}&\textbf{593.2}&\textbf{89.5}\\
  \hline
  Naive Complete&80\%&343.8&145.8\\
  \hline
  Many Iterations Ring&25\%&223.8&67.1\\
  \hline
  Many Iterations Random&25\%&98.6&11.0\\
  \hline
  \end{tabular}
\end{table}

\begin{table}
  \caption{Summary of results for function Sphere with 50 dimensions,
  measuring number of time steps to reach a value of 1e-06}
  \label{tab:sphere-50}
  \centering
  \begin{tabular}{|l|r|r|r|}
  \hline
  Algorithm&\% Complete&Mean&St. Dev.\\
  \hline
  \hline
  Naive Ring&0\%&N/A&N/A\\
  \hline
  Naive Random&30\%&986.8&8.2\\
  \hline
  Naive Complete&100\%&458.4&12.3\\
  \hline
  Many Iterations Ring&100\%&288.7&3.9\\
  \hline
  \textbf{Many Iterations Random}&\textbf{100\%}&\textbf{248.6}&\textbf{5.1}\\
  \hline
  \end{tabular}
\end{table}

\begin{table}
  \caption{Summary of results for function Schwefel with 50 dimensions,
  measuring number of time steps to reach a value of 80}
  \label{tab:schwefel-50}
  \centering
  \begin{tabular}{|l|r|r|r|}
  \hline
  Algorithm&\% Complete&Mean&St. Dev.\\
  \hline
  \hline
  Naive Ring&0\%&N/A&N/A\\
  \hline
  Naive Random&70\%&871.6&57.1\\
  \hline
  Naive Complete&100\%&599.0&85.6\\
  \hline
  \textbf{Many Iterations Ring}&\textbf{100\%}&\textbf{238.6}&\textbf{34.2}\\
  \hline
  \textbf{Many Iterations Random}&\textbf{100\%}&\textbf{242.9}&\textbf{95.4}\\
  \hline
  \end{tabular}
\end{table}

>>>>

In our approach, we refactor PSO such that the computation needed for iteration
$t+1$ can be done concurrently with the computation needed for iteration $t$.
Thus we can perform two iterations of PSO at once.  Even with some amount of
wasted computation, we show that this approach to parallelization in PSO often
outperforms the standard parallelization of simply adding particles to the
swarm.  SEPSO produces results that are exactly equivalent to PSO; this is not
a new algorithm or variant, only a new method of parallelization.

@@

Furthermore, we show that if we relax the requirements of the algorithm, no
longer demanding that it strictly reproduce the exact behavior of standard PSO,
we can introduce new speculative techniques that often out-perform both
standard parallelizations of PSO and SEPSO.  These relaxations make better use
of the information obtained from the extra exploration made by the speculative
function evaluations.  We also explore the idea that, like branch prediction in
processors, we need not speculatively evaluate \emph{all} possible future
positions, we can accelerate the algorithm even if we are just \emph{likely} to
have guessed right.  By pruning the speculation to just paths that are
statistically likely to reproduce the paths that are equivalent to PSO we can
increase the swarm size without increasing the number of speculative
evaluations.  We also consider several recovery strategies for cases where the
pruned set of speculative evaluations does not contain the evaluation that
standard PSO would have done.  A further improvement we explore is speculating
several iterations ahead instead of just one, which is made possible by pruning
the number of speculative evaluations.

@@

While the idea of speculative evaluation in particle swarm optimization is
intriguing, the results in \sec{results} show that in many cases it requires
too many extra evaluations to produce competitive results.  On most functions,
the larger swarm size that can be used in standard parallelizations leads to
better performance than is found by trying to do speculative evaluation to
exactly reproduce PSO.  However, if we keep the idea of speculative evaluation
while relaxing the requirement of exactly reproducing the behavior of the
original PSO algorithm, we see some impressive results.

@@

Because the problem facing speculative evaluation is having too many possible
evaluations to perform, a natural step to take is to eliminate some of the
evaluations.  If we could reliably predict which branch were going to be taken,
we could limit ourselves to one speculative evaluation per particle instead of
$2n+1$.  With a fixed number of processors, this would allow us to greatly
increase the swarm size relative to that needed in the original speculative
algorithm (e.g., with 120 processors, a standard parallelization has a swarm of
size 120, complete speculative evaluation has a swarm of size 15, and pruning
the evaluations to only one per particle allows a swarm of size 60).  As not
all of the branches are evaluated in any given iteration, we call this
technique pruning.

@@

We ran experiments to compare our speculative parallelization of PSO to the
standard parallelization.  At each iteration of the algorithms, we use one
processor to perform one function evaluation for one particle, be it
speculative or non-speculative.  The speculative version actually performs two
iterations of PSO at each ``iteration,'' so we instead call each ``iteration''
a ``time step.''  As noted in the introduction, this is a reasonable
architecture to use if the function being optimized takes longer than around
100 milliseconds to run in our implementation.

@@

Using the same number of processors for each approach requires that our
speculative parallelizations have a smaller swarm size than the standard
parallelization.  For each of the topologies we used, a particle has three
neighbors including itself.  As shown in Table~\ref{tab:evals}, this results in
$7$ speculative evaluations per particle.  With one evaluation needed for the
original, non-speculative particle, we have $8p$ evaluations for every two
iterations, where $p$ is the number of particles in the speculative swarm.  The
extra evaluations required in our speculative approach would instead be used to
evaluate particles in standard parallelizations, so we compare swarms of size
$p$ in speculative evaluation with swarms of size $8p$ in standard approaches.

@@

That being said, given the time required for the evaluation of the objective
function and the communication overhead per iteration for a specific
implementation, the actual running time can be recovered from all of our
results.  Simply multiply the number of time steps by the sum of the function
evaluation time and the overhead.  We wish to stress that the ``time steps'' we
report are proportional to wall-clock time, given the architecture we have
assumed.

@@

The parallelization techniques we compare are the standard parallelization
(here labeled Standard simply for ease of reference), our original speculative
approach (labeled SEPSO), and the four relaxations of SEPSO discussed in
\sec{relax}.  In presenting our results, we call the approach developed in
\sec{pickbest} Pick Best.  The methods described in \sec{pruning} through
\sec{wrong} are called Pick Best Pruned and Social Promotion Pruned, and the
method in \sec{manyiters} we call Many Iterations.

@@

Because pruning only requires two speculative evaluations per particle (along
with evaluating the original particle), we can use swarms of size
$\frac{1}{3}p$ to compare to a swarm of size $p$ with the standard
parallelization, instead of $\frac{1}{8}p$ with other techniques.

@@

Where the Complete topology would normally be preferred, we use a Random
topology in SEPSO and Pick Best, as Complete leads to an explosion in the
number of speculative evaluations (as noted, the other speculative techniques
can still use a Complete topology; we often report results on both topologies
for those methods).  If speculative evaluation were not being performed, it is
possible that the Complete topology would be used.  However, the Complete
topology also requires a very large amount of interprocessor communication in
distributed PSO, so it is still quite possible that Random would be used even
with standard parallelizations~\cite{mcnabb-2009-large-particle-swarms}.  But,
to be fair in our comparisons, we compare to the standard parallelization using
both the Random topology and the Complete topology (labeled PSO Random and PSO
Complete in our results).  Again, the amount of communication overhead is
heavily dependent on implementation details which are not the focus of this
paper.  A practitioner using a particular implementation can compare results
for whichever topology is most practical given the specific implementation.

@@

Rastrigin is a multi-modal function that is best optimized with a large,
Complete swarm.  It has been shown that with Rastrigin, the more particles
there are in the swarm, the lower function value it finds, up to at least 4000
particles~\citep{mcnabb-2009-large-particle-swarms}.  Smaller swarms get caught
in local optima.  Because our speculative algorithms require significantly
smaller swarm sizes, we would expect to not perform very well on functions such
as Rastrigin.  Our experiments show our intuition to be correct.  In this
experiment we used 240 processors, so SEPSO and Pick Best each had 30
particles, and the Standard algorithms had 240 particles.  As expected, SEPSO
and Pick Best converge quickly to worse local optima than Standard does.
\fig{rastrigin} shows the results graphically, and \tabref{rastrigin-20} shows
results for all of the methods we tried.

\begin{table}
  \caption{Summary of results for function Rastrigin with 20 dimensions,
  measuring number of time steps to reach a value of 20}
  \label{tab:rastrigin-20}
  \centering
  \begin{tabular}{|l|r|r|r|}
  \hline
  Algorithm&\% Complete&Mean&St. Dev.\\
  \hline
  \hline
  Standard Ring&25\%&743.4&212.0\\
  \hline
  Standard Random&100\%&372.1&86.6\\
  \hline
  \textbf{Standard Complete}&\textbf{95\%}&\textbf{273.1}&\textbf{99.7}\\
  \hline
  SEPSO Ring&5\%&471.0&0.0\\
  \hline
  SEPSO Random&15\%&208.0&17.0\\
  \hline
  Pick Best Ring&35\%&489.0&152.0\\
  \hline
  Pick Best Random&15\%&136.0&40.1\\
  \hline
  Pick Best Pruned Ring&95\%&545.9&208.5\\
  \hline
  Pick Best Pruned Random&10\%&95.0&11.0\\
  \hline
  Pick Best Pruned Complete&5\%&59.0&0.0\\
  \hline
  Social Promotion Pruned Ring&5\%&580.0&0.0\\
  \hline
  Social Promotion Pruned Random&80\%&278.2&136.3\\
  \hline
  Social Promotion Pruned Complete&50\%&200.5&130.2\\
  \hline
  Many Iterations Ring&10\%&117.0&25.0\\
  \hline
  Many Iterations Random&5\%&57.0&0.0\\
  \hline
  Many Iterations Complete&0\%&N/A&N/A\\
  \hline
  \end{tabular}
\end{table}

@@

We ran 50 trials of each experiment with Griewank, so that the curves are more
smooth.  We show results in \fig{basic-griewank1} for swarms of size 30 and 240
using the Ring topology.  One can see in the figure that the SEPSO approach
finds the optimum faster than the Standard approach.  However, because the
swarm size is so small, SEPSO gets stuck a little less than half of the time.

@@

Because the speculative algorithm got stuck with such a small swarm size, we
ran a set of experiments on Griewank with 800 processors, giving swarms of size
100 and 800 instead of 30 and 240.  \fig{basic-griewank2} shows the results.
\fig{basic-griewank2} is very similar to \fig{basic-griewank1}, except that
speculative evaluation finds the global optimum essentially 100\% of the time.
Thus with a few more processors, our algorithm finds the optimum on average
twice as fast as the Standard approach without a significant loss of accuracy.

@@

With Griewank, the premature convergence problems inherent in picking the best
child are exacerbated when speculating several iterations ahead.  When Many
Iterations finds the optimum, it finds it quicker than any other method we
tried, on average four times faster than Standard.  However, it also gets stuck
and fails to find the optimum more than any other method.  The results are
shown in \fig{griewank-manyiters}.   \fig{griewank-manyiters} is also
interesting in that it highlights the trade-off between accuracy and speed in
the various approaches at this swarm size.  The faster the approach finds the
optimum, the less likely it is to be successful.

@@

\begin{table}
  \caption{Summary of results for function Sphere with 20 dimensions,
  measuring number of time steps to reach a value of 1e-35}
  \label{tab:sphere-20}
  \centering
  \begin{tabular}{|l|r|r|r|}
  \hline
  Algorithm&\% Complete&Mean&St. Dev.\\
  \hline
  \hline
  Standard Ring&0\%&N/A&N/A\\
  \hline
  Standard Random&0\%&N/A&N/A\\
  \hline
  Standard Complete&100\%&931.0&19.1\\
  \hline
  SEPSO Ring&0\%&N/A&N/A\\
  \hline
  SEPSO Random&85\%&972.8&15.8\\
  \hline
  Pick Best Ring&100\%&917.4&21.0\\
  \hline
  Pick Best Random&100\%&768.1&14.3\\
  \hline
  Pick Best Pruned Ring&100\%&967.8&9.9\\
  \hline
  Pick Best Pruned Random&100\%&693.5&10.4\\
  \hline
  \textbf{Pick Best Pruned Complete}&\textbf{100\%}&\textbf{389.6}&
    \textbf{9.6}\\
  \hline
  Social Promotion Pruned Ring&0\%&N/A&N/A\\
  \hline
  Social Promotion Pruned Random&100\%&971.1&10.6\\
  \hline
  Social Promotion Pruned Complete&100\%&777.8&16.7\\
  \hline
  Many Iterations Ring&100\%&575.2&9.8\\
  \hline
  Many Iterations Random&100\%&442.4&7.1\\
  \hline
  Many Iterations Complete&0\%&N/A&N/A\\
  \hline
  \end{tabular}
\end{table}

@@

Schwefel 2.21 is a function similar to Sphere, but Schwefel 2.21 is benefited
more by larger swarms than Sphere is.  Thus our speculative algorithms often
fail outperform Standard with 240 processors because the simple speculative
methods only have 30 particles.  As we will show later, when we use 800
processors at 50 dimensions, our methods perform much better.  However, the
pruned versions of our methods have 80 particles and thus are able to
outperform Standard even with only 240 processors.  Because the graph of
Schwefel 2.21 looks very similar to that of Sphere, we present all of our
results in \tabref{schwefel-20}.

\begin{table}
  \caption{Summary of results for function Schwefel with 20 dimensions,
  measuring number of time steps to reach a value of 1e-06}
  \label{tab:schwefel-20}
  \centering
  \begin{tabular}{|l|r|r|r|}
  \hline
  Algorithm&\% Complete&Mean&St. Dev.\\
  \hline
  \hline
  Standard Ring&0\%&N/A&N/A\\
  \hline
  Standard Random&0\%&N/A&N/A\\
  \hline
  Standard Complete&100\%&837.2&44.8\\
  \hline
  SEPSO Random&0\%&N/A&N/A\\
  \hline
  Pick Best Random&5\%&938.0&0.0\\
  \hline
  \textbf{Pick Best Pruned Random}&\textbf{100\%}&\textbf{639.9}&\textbf{29.4}\\
  \hline
  \textbf{Pick Best Pruned Complete}&\textbf{100\%}&\textbf{597.7}&
	\textbf{77.9}\\
  \hline
  Social Promotion Pruned Random&75\%&938.0&33.0\\
  \hline
  Social Promotion Pruned Complete&100\%&815.2&43.5\\
  \hline
  Many Iterations Random&100\%&783.5&94.5\\
  \hline
  Many Iterations Complete&0\%&N/A&N/A\\
  \hline
  \end{tabular}
\end{table}

@@

Bohachevsky is a unimodal function best optimized with a Complete swarm.  It is
similar to Griewank in that there is a global optimum with a value of 0, and
the swarm sometimes finds it and sometimes does not.  Thus we present a graph
similar to those of Griewank, because ``average value'' graphs have a
misleading flat line.  In \fig{bohachevsky} we show a plot with a few pertinent
methods, while \tabref{bohachevsky-20} shows all of the results.  All of our
speculative approaches found the optimum much quicker than Standard with a
Random topology.  However, SEPSO was slower than Standard Complete and got
stuck 25\% of the time.  Pick Best, Pick Best Pruned, and Many Iterations all
outperformed Standard Complete, with Many Iterations finding the optimum about
twice as fast.

@@

\begin{table}
  \caption{Summary of results for function Bohachevsky with 20 dimensions,
  measuring number of time steps to reach a value of 1e-06.}
  \label{tab:bohachevsky-20}
  \centering
  \begin{tabular}{|l|r|r|r|}
  \hline
  Algorithm&\% Complete&Mean&St. Dev.\\
  \hline
  \hline
  Standard Random&100\%&472.3&6.6\\
  \hline
  Standard Complete&100\%&238.2&11.2\\
  \hline
  SEPSO Random&75\%&261.6&9.2\\
  \hline
  Pick Best Random&100\%&211.8&9.2\\
  \hline
  Pick Best Pruned Random&100\%&189.7&3.4\\
  \hline
  Pick Best Pruned Complete&FIXME!&FIXME!&FIXME!\\
  \hline
  Social Promotion Pruned Random&100\%&260.3&6.5\\
  \hline
  Social Promotion Pruned Complete&FIXME!&FIXME!&FIXME!\\
  \hline
  \textbf{Many Iterations Random}&\textbf{94\%}&\textbf{118.2}&\textbf{2.6}\\
  \hline
  Many Iterations Complete&0\%&N/A&N/A\\
  \hline
  \end{tabular}
\end{table}

@@

Given our observation from the 20 dimensional benchmark functions that Many
Iterations most often outperforms other speculative approaches, we only present
results for Many Iterations and Standard for the 50 and 500 dimensional
variants.  For these experiments we used 800 processors, as the problems are
more difficult and speculative approaches perform best when many processors are
available, as we showed in \sec{griewank-20}.  Many Iterations Complete showed
the same premature convergence that was observed with the 20 dimensional
benchmark functions, so we do not include those results in our tables; the
success rate was 0\% for all functions.

Tables~\ref{tab:rastrigin-50} through~\ref{tab:bohachevsky-50} present
summaries of our results for each of the benchmark functions. As with the 20
dimensional variants, our speculative methods did not perform well on
Rastrigin, though for all other benchmarks Many Iterations significantly
outperformed Standard.  In each table, we bold the best method, meaning that it
has at least a 90\% success rate and its average time to completion is faster
than all other methods that have at least a 90\% success rate.  For example,
Many Iterations Random, when it is successful, achieves the required value 6
times faster than Standard Random.  However, because Many Iterations Random
only has a 25\% success rate while Standard Random has a 100\% success rate,
Standard Random is bolded.

@@

A summary of the results is shown in \tabref{50dims-summary}.  For functions
where Many Iterations outperformed Standard, we report the average speed up.
On all functions except Rastrigin, our methods showed an average speed up of
from two to six times compared to previous methods.

\begin{table}
  \caption{Summary of results for function Rastrigin with 50 dimensions,
  measuring number of time steps to reach a value of 100}
  \label{tab:rastrigin-50}
  \centering
  \begin{tabular}{|l|r|r|r|}
  \hline
  Algorithm&\% Complete&Mean&St. Dev.\\
  \hline
  \hline
  Standard Ring&5\%&776.0&0.0\\
  \hline
  \textbf{Standard Random}&\textbf{100\%}&\textbf{593.2}&\textbf{89.5}\\
  \hline
  Standard Complete&80\%&343.8&145.8\\
  \hline
  Many Iterations Ring&25\%&223.8&67.1\\
  \hline
  Many Iterations Random&25\%&98.6&11.0\\
  \hline
  \end{tabular}
\end{table}

\begin{table}
  \caption{Summary of results for function Sphere with 50 dimensions,
  measuring number of time steps to reach a value of 1e-06}
  \label{tab:sphere-50}
  \centering
  \begin{tabular}{|l|r|r|r|}
  \hline
  Algorithm&\% Complete&Mean&St. Dev.\\
  \hline
  \hline
  Standard Ring&0\%&N/A&N/A\\
  \hline
  Standard Random&30\%&986.8&8.2\\
  \hline
  Standard Complete&100\%&458.4&12.3\\
  \hline
  Many Iterations Ring&100\%&288.7&3.9\\
  \hline
  \textbf{Many Iterations Random}&\textbf{100\%}&\textbf{248.6}&\textbf{5.1}\\
  \hline
  \end{tabular}
\end{table}

\begin{table}
  \caption{Summary of results for function Schwefel with 50 dimensions,
  measuring number of time steps to reach a value of 80}
  \label{tab:schwefel-50}
  \centering
  \begin{tabular}{|l|r|r|r|}
  \hline
  Algorithm&\% Complete&Mean&St. Dev.\\
  \hline
  \hline
  Standard Ring&0\%&N/A&N/A\\
  \hline
  Standard Random&70\%&871.6&57.1\\
  \hline
  Standard Complete&100\%&599.0&85.6\\
  \hline
  \textbf{Many Iterations Ring}&\textbf{100\%}&\textbf{238.6}&\textbf{34.2}\\
  \hline
  \textbf{Many Iterations Random}&\textbf{100\%}&\textbf{242.9}&\textbf{95.4}\\
  \hline
  \end{tabular}
\end{table}
